---
title: "Data Wrangling Intro"
author: "Desik Somasundaram"
date: "8/16/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(palmerpenguins)
```

## Warm-up:

Look at 'us_rent_income'. Make a tidy version. Get back to the original structure.

```{r}
```


```{r}
us_rent_income_long <- us_rent_income %>% 
  select(-moe) %>% 
  pivot_wider(names_from = variable, values_from = estimate)

```

```{r}
us_rent_income_long_original <- us_rent_income_long %>%
  pivot_longer(cols = income:rent, names_to = "variable", values_to = "estimate" )
  
```

Only keep observations where the island is Biscoe

```{r}
ex_1 <- filter(.data = penguins, island == "Biscoe")
unique(ex_1$island)

```
Only keep observations if the penguin species is Chinstrap AND the year is 2008

```{r}
ex_2 <- penguins %>% 
  filter(species == "Chinstrap", year == 2008)
```

Keep observations if the penguins species is NOT "Gentoo"
```{r}
ex_3 <- penguins %>% 
  filter(species != "Gentoo")
```

Create a subset from 'penguins' that only contains observations for Adelie penguins with a flipper length greater than of equal to 185 mm

```{r}
ex_4 <- penguins %>% 
  filter(species == "Adelie", flipper_length_mm >= 185)

```

Use the %in% operator to look for *any* matches within a vector,  and return TRUE if the value does exist for our row.

We want to keep observations where the island is Dream or Torgersen.

```{r}
ex_5 <- penguins %>% 
  filter(island %in% c("Dream", "Torgersen"))


```

Select() to select or exclude COLUMNS

5. Keep a single column by name

Only want to keep the species column

```{r}
ex_6 <- penguins %>%  select(species) 

ex_7 <- select(.data = penguins, species)
```

Keep multiple columns by name

Create a subset that contains species, bill_length_mm through flipper_length_mm, and year

```{r}
ex_8 <- penguins %>% 
  select(species, bill_length_mm:flipper_length_mm, year)
```

```{r}
ex_9 <- penguins %>% 
  select(-4)
```

Piped sequence where you first only keep columns for species, body_mass_g, and island, and second: filter to only include observations if body_mass_g is greater than 3600

```{r}
ex_10 <- penguins %>% 
  select(species, body_mass_g, island) %>% 
  filter(body_mass_g > 3600)

```

Mutate() to add or update existing columns, while keeping the ones already there

Add a column that converts body mass from g to kg

```{r}
ex_11 <- penguins %>% 
  mutate(body_mass_kg = body_mass_g / 1000) %>% 
  mutate(flipper_length_m = flipper_length_mm / 1000)

```

Mutate to update an existing column

```{r}
ex_12 <- penguins %>% 
  mutate(island = str_replace(island, pattern = "sen", replacement = "HOORAY!"))

```

Update the class of the bill depth column to character

```{r}
ex_13 <- penguins %>% 
  mutate(bill_depth_mm = as.character(bill_depth_mm)) 

```

case_when(): your favorite if-else

Create a new column that contains "small" if penguin flipper length is less than 185 mm, 
"medium" otherwise
```{r}
ex_14 <- penguins %>% 
  mutate(binned_size = case_when(
    flipper_length_mm < 185 ~ "small", flipper_length_mm >= 185 ~ "medium"
  ))

```

new column "bird_bill_bin" that contains:

-If bill length < 38 mm: "short bill"
-If bill length >= 38mm & < 43 mm: "medium bill"
-If bill length >= 43: "long bill"

```{r}
ex_15 <- penguins %>% 
  mutate(bird_bill_bin = case_when(
    bill_length_mm < 38 ~ "short bill", 
    bill_length_mm >= 38 & bill_length_mm < 43 ~ "medium bill", 
    bill_length_mm > 43 ~ "long bill"
  ))

```

group-by() %>% summarize() to find summary values by group, and return them in a nice summary table

```{r}
mean(penguins$bill_depth_mm, na.rm = TRUE)
```

```{r}
ex_16 <- penguins %>% 
  group_by(species, island) %>% 
  summarize(mean_flipper_length = mean(flipper_length_mm, na.rm = TRUE), sd_flipper_length = sd(flipper_length_mm, na.rm = TRUE))

ex_16

```

In piped sequence:

-Only keep penguins that have a body mass > 3500 g
-Only keep columns for species, flipper_length_mm, and sex
-Create a summary table that contains the mean, sd, max, min value of flipper
length by species and sex (as the grouping variables)

```{r}
prac <- penguins %>% 
  filter(body_mass_g > 3500) %>% 
  select(species, flipper_length_mm, sex) %>% 
  group_by(species, sex) %>% 
  summarize(mean_flipper_length = mean(flipper_length_mm, na.rm = TRUE), sd_flipper_length = sd(flipper_length_mm, na.rm = TRUE), max_flipper_length = max(flipper_length_mm, na.rm = TRUE), min_flipper_length = min(flipper_length_mm, na.rm = TRUE) )

```

